---
title: "R-Ladies Abuja - Data Manipulation with data.table in R - hands on session"
author: "Njoki Njuki Lucy"
date: "19/03/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      eval = FALSE)
```

## Hands-on session

1. Load the starwars data from the dplyr package and convert to data.table.

```{r load_dplyr_package}
library(dplyr)
library(data.table)
```

```{r load_starwars_dataframe}
#let's load the data
starwars

#check the info about the data variables
?starwars
```

```{r convert_DF_to_DT}
#convert DF to DT
(starwarsDT <- as.data.table(starwars))
starwarsDT

class(starwarsDT)
```

```{r str_starwarsDT}
#Structure of the DT
str(starwarsDT)

View(starwarsDT)
```


  1. Filter the observations which have droid and human for species.
  
```{r filter_rows}
# filter DT to have only Droid and Human as species
starwarsDT[species %in% c("Droid", "Human")]
```
  
  2. Compute the average weight per species
  
```{r create_new_cols}
#find mean by species
starwarsDT[,.(meanWeight = mean(mass, na.rm = T)),
           by = .(species)]
```
  
  3. Consider the females with information concerning height and mass.
  
```{r filter_data_with_females}
# filter DT to have only females
(FstarwarsDT <- starwarsDT[sex == "female"])
```

  1. Compute the BMI.
  
  $BMI = Weight(Kg)/{Height(m)}^{2}$
  
```{r calc_BMI}
#W/o the existing cols .()/list()
FstarwarsDT[,.(bmi = mass/(height/100)^2)]

#W the other cols
FstarwarsDT[, "bmi" := mass/(height/100)^2]
FstarwarsDT
```
  
  
2. Load the band_members and band_instruments data from the dplyr package and convert to data.table.

```{r load_band_tibbles}
#load band_members; band_instruments
band_members

band_instruments
```

```{r convert_band_tiblles_to_DT}
#convert the tibbles to DT
(band_membersDT <- as.data.table(band_members))

(band_instrumentsDT <- as.data.table(band_instruments))
```

  1. Combine all information from the members and the instruments.
  
Full join:
  
```{r band_members_instrumentsDT}
#full join band_instruments and band_members
(band_members_instrumentsDT <- merge(x = band_membersDT, y = band_instrumentsDT, by = "name", all = TRUE))
```

## References

1. Hadley Wickham, Romain François, Lionel Henry and Kirill Müller (2022). dplyr: A Grammar of Data Manipulation. R package version 1.0.8. https://CRAN.R-project.org/package=dplyr
2. Matt Dowle and Arun Srinivasan (2021). data.table: Extension of `data.frame`. R package version 1.14.2. https://CRAN.R-project.org/package=data.table